---
title: "qcdrift: Automated QC Drift Correction for Metabolomics"
format: gfm
---

```{r}
#| echo: false
#| setup: true
#| message: false

# load the current working version of the package to render the documentation and run the example pipeline.
devtools::load_all()
```

The `qcdrift` package is an R-based analytical framework designed to automate the
detection, correction, and validation of instrumental drift in high-dimensional 
mass spectrometry data. It replaces labor-intensice manual spreadsheet workflows
with a reproducible, modular pipeline. 


## Quick Start 

If you have never used R or RStudio before, follow these steps to get up and running. 

### Preparation (one-time-setup)

Before running the analysis, you must install the tools R needs to create your 
graphs.

1. Open RStudio
2. Copy and paste the following line into the Console (the bottom-left window)
and press Enter: 

```{r}
#| eval: false
install.packages(c("devtools", "dplyr", "ggplot2", "patchwork", "ggrepel", "GGally","readxl","tidyr"))
```

### Installation

`qcdrift` can be installed direclty from GitHub using `devtools`:

```{r}
#| eval: false
devtools::install_github('Hood-BIFX/qcdrift')
```


## User Guide & Functional Walkthrough

The pipeline is engineered with a modular architecture, allowing users to call individual functions for specific tasks or use the `process_runs()` for a complete end-to-end analysis. 

### Automated Workflow

Once the package is installed and loaded, you can run the entire pipeline with a single command. For this example, we'll use the provided example dataset located at `inst/extdata/RawMassSpec.xlsx`.

```{r}
#| eval: false

library(qcdrift)

results <- system.file('extdata/RawMassSpec.xlsx', package = 'qcdrift') |>
  process_runs()
```

The resulting `results` object is a comprehensive list containing the cleaned and corrected data, as well as all generated diagnostic plots. A multi-page PDF report can also be saved in your working directory, summarizing the key findings as follows:

```{r}
#| eval: false

# We seem to have lost this at some point, so this is a placeholder for the final report rendering function.
render_report(results, output_file = "Final_QC_Report.pdf")
```

### Individual Functional Modules 

The pipeline is built on modular functions that can be called independently. We will walk through the key functions in this section, using the example dataset to illustrate their use and outputs.

#### Data Import and Cleaning

`read_and_clean_data()` standardizes the raw Excel into a "tidy" long-format data frame.

* Automated Parsing: It extracts sample names and numerical injection orders
from the file header.
* QC Identification: It uses regular expression to classify samples based on the `qc_starts_with` parameter, which defaults to "QC".

```{r}
raw_data <- system.file('extdata/RawMassSpec.xlsx', package = 'qcdrift') |>
  read_and_clean_data()
```

#### Drift Correction

`qc_drift_correction()` applies the `corret_linear()` function (piecewise linear interpolation between QC samples) to correct for QC drift (other options may be added in the future). The function expects the input data frame to have the following columns, which are automatically generated by `read_and_clean_data()`:

* `io`: The injection order (numerical sequence of samples)
* `abundance`: The raw intensity values for each metabolite
* `qc`: A binary vector indicating which samples are QCs (1 for QC, 0 for non-QC)
* `metabolites`: The name or ID of the molecule being measured

```{r}
corrected_data <- qc_drift_correction(raw_data)
```

#### Normalization

`normalize_data()` applies Total Sum Normalization (TSN) to account for injection-level variability, such as differences in sample volume or dilution. Alternately, specifying `method = 'auto'` will applly autoscaling (normalization to mean=0 and sd=1) if desired.

```{r}
normalized_data <- normalize_data(corrected_data)
```

### Visualization

The package includes several `ggplot2`- based functions to validate data integrity.

#### Principal Component Analysis: `generate_pca_plot()`
* Function: Visualizes the global variance structure 
* Interpretation: Tight clustering of QC sample (blue diamonds) in the "Corrected"
plot indicates successful removal of technical noise. 

#### Precision Tracking `generate_cv_plot()`

* Function: Creates a waterfall plot of Coefficient of Variation (CV) percentages
* Interpretation: A horizontal dashed line at 20% marks the acceptable industry
threshold for reproducibility. 

#### Distribution Analysis: `generate_qc_violin()`

* Function: Display log10-transformed abundance distributions across the run.
* Interpretation: Consistent median values and interquartile range across samples
indicate successful Total Sum Normalization (TSN). 

#### Heatmap Visualization: `generate_qc_heatmap()`

* Function: Displays Z-score normalized intensities across injection orders. 
* Interpretation: The removal of "stripping" (verrtical color gradients) confrims
that temporal decay has been eliminated. 
